FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-03-05 ess_marmo.tex
Minor changes for submission. In-file bibliography.

%% Choose from this list of Journals:
%
% JGR-Atmospheres
% JGR-Biogeosciences
% JGR-Earth Surface
% JGR-Oceans
% JGR-Planets
% JGR-Solid Earth
% JGR-Space Physics
% Global Biochemical Cycles
% Geophysical Research Letters
% Paleoceanography
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resource Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science

@@

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)

@@

% Example: \correspondingauthor{First and Last Name}{email@address.edu}

% Authors are individuals who have significantly contributed to the
% research and preparation of the article. Group authors are allowed, if
% each author in the group is separately identified in an appendix.)

% \affiliation{1}{First Affiliation}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

%% Keypoints, final entry on title page.
% Example: 
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 100 characters or less with no special
% characters or punctuation 
% \end{keypoints}

%% \begin{abstract} begins second page

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Track Changes:
% To add words, \added{<word added>}
% To delete words, \deleted{<word deleted>}
% To replace words, \replace{<word to be replaced>}{<replacement word>}
% To explain why change was made: \explain{<explanation>}

% When final version, \listofchanges will not produce anything,
% \added{} word will be printed, \deleted{} will take away the word,
% \replaced{}{} will print only the 2nd argument.
% \explain will not print anything.

% Optional argument:
% You can also add additional information to be printed with the list
% of changes, to indicate the initials of the person changing the text,
% and the time and/or date of the change, or any other comment by using
% the optional [] argument:
% \added[AH, 3:30pm, Feb 18, 2016]{added term}
% will yield 
% [AH, 3:30pm, Feb 18, 2016] added term on page...
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
Planetary science encompasses a broad number of research fields and
brings together several research communities (geologists, astronomers,
physicists, geochemists, etc.). Planetary missions produce an impressively growing 
amount of diverse data requiring an evolution from a mostly manual-based visual 
analysis to a more automated and quantitative analysis.
Interoperability, openness of data formats, and shared processing techniques are 
a necessity to efficiently extract scientific information from the data and 
to guarantee the reproducibility of the scientific results.
Unfortunately, the technologies and data formats used by researchers for
planetary surface studies and the field of astronomy 
diverged as these related, but almost completely isolated domains evolved.  
% * <thare@usgs.gov> 2018-02-13T00:44:29.292Z:
% 
% > Unfortunately, the technologies 
% > and data formats used by researchers for planetary surface studies and the field of astronomy 
% > diverged as these related, but almost completely isolated domains evolved.  
% 
% Is this too strong a statement?
% 
% ^.
In this paper we will describe how a small addition to the Flexible Image Transport
System (FITS) standard, widely used in the astronomy investigations, will allow FITS
to be more easily used in planetary surface investigations.
We will also show how FITS metadata can easily be transformed
in the PDS4 metadata archival model and lastly provide example implementations.
More than imposing a formal data model, a FITS description for planetary data aims
to simplify sharing data across the planetary and astronomy domains and promoting
interoperability from raw data formatting to final visualization.
\end{abstract}

@@

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition,
transfer and archiving of telescope data by astronomical observatories.
It had been used for data exchange and archiving from several orbital telescopes.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications so that FITS files can be embedded in PDS datasets.
Finally, it is supported by a large number of open libraries
and software tools.

@@

\begin{table}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: the corresponding PDS4 terminology is
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator: those three projection are not used for archival or analysis.}
\end{tabular}
\label{table:proj}
\end{table}

Once reprojected in one of the standard projections listed in Table \ref{table:proj}
Planetary images or maps can be described using the WCS scheme
\citep{greisencalaII,calagreisenI}.
In WCS context pixels are first translated in an intermediate pixel coordinate system
via linear transformations, then mapped on the grid specified by the \texttt{CTYPE}
keyword, containing the projection code and the coordinate system.
% * <chiara.marmo@u-psud.fr> 2018-02-22T21:31:39.056Z:
% 
% Example or scheme needed in order to better understand the coordinate transformation flow
% 
% ^.

The coordinate system must be defined with respect to a specific inertial reference frame.
% * <chiara.marmo@u-psud.fr> 2018-02-26T11:33:56.098Z:
% 
% Please be more specific about the inertial frame
% 
% ^.
Multiple coordinate system descriptions are allowed in the same header: this is a
valuable property for planetary surface imagery, as angular (`deg') and linear (`m')
description of the surface are often needed at the same time.
Further below 3D keyword will be proposed for FITS metadata, in order to standardize
the conversion between angular and linear coordinates: alternate description will still
be important in our framework as standard libraries (e.g.
wcslib\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/} or
astropy\footnote{http://www.astropy.org/} \citet{astropy}) and some FITS visualization
software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them, making linear
coordinate tracking more efficient.

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in UTC but other systems are accepted as long as they
are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\textit{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UTC.
The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP
scheme, they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}

@@

%
%\item{\texttt{AUTHOR}}

@@

In addition to that, a QGIS plugin has been developed\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
\citep{Minin} in order to download and display files exposed via EPN-TAP servers.

@@

\subsection{Aladin}
Aladin is both a sky atlas and a portal to access data -- mainly images and
catalogues -- available through the VO.
It is able to load FITS files and to display images taking into account various
WCS projections.
Like DS9, Aladin is popular and broadly used in the part of the astronomical
community manipulating images.
So far it is also the reference generator and visualizer of Hierarchical
Progressive Survey (HiPS).
HiPS\footnote{http://www.ivoa.net/documents/HiPS/} is a recommendation endorsed
by the International Virtual Observatory Alliance (IVOA).
It describes how to access to sky survey data stored at various spatial
resolutions, offering a progressive view of possibly very large datasets.
%The purpose is to allows users to explore all-sky surveys without having to
%explicitly perform any query: the most important information is visible at low
%spatial resolutions while finer details appear by zooming in.
%In the framework of the VESPA project, the idea of using Aladin to display
%planetary data, especially planetary HiPS, has emerged.
HiPS are usually generated from a collection of WCS tagged FITS files.
%As an alternative, Aladin can use JPEG files together with additional metadata
%files.
%A metadata file -- one by JPEG file -- basically contains in a text format the
%WCS part of a FITS header.
Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
been updated to support planetary conventions (see Figure \ref{fig:aladin}).
%Then, planetary missions usually use different projections when providing
%images of planets polar caps and equatorial region.
%Often the former are in the stereographic (STG) projection while the later
%is in the Plate Carr\'ee (CAR) projection. 
%This specificity has been taken into account into the HiPS generator.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendix goes here
%
% \appendix resets counters and redefines section heads
% but doesn't print anything.
% After typing \appendix
%
%\section{Here Is Appendix Title}
% will show
% A: Here Is Appendix Title
%
%\appendix
%\section{Here Is Appendix Title}
%This is an Appendix section.

%\begin{linenomath*}
%\begin{equation}asdf\end{equation}
%\end{linenomath*}

\appendix
\section{Planetary Header example}
%An example of planetary header goes here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Optional Glossary, Notation or Acronym section goes here:
%
%%%%%%%%%%%%%%
% Glossary is only allowed in Reviews of Geophysics
%\begin{glossary}
%\term{Term}
% Term Definition here
%\term{Term}
% Term Definition here
%\term{Term}
% Term Definition here
%\end{glossary}

@@

\bibliography{ess_marmo}

@@

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%% ------------------------------------------------------------------------ %%
%
%  IN-TEXT LISTS
%
%% ------------------------------------------------------------------------ %%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%% ------------------------------------------------------------------------ %%
%
%  EQUATIONS
%
%% ------------------------------------------------------------------------ %%

% Single-line equations are centered.
% Equation arrays will appear left-aligned.

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

 Math coded inside \begin{equation} and \end{equation} will
 be automatically numbered, e.g.,:
 \begin{equation}
 x^2=y^2 + z^2
 \end{equation}

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%% ------------------------------------------------------------------------ %%
%
%  EQUATION NUMBERING: COUNTER
%
%% ------------------------------------------------------------------------ %%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

% If you are using line numbers, remember to surround
% equations with \begin{linenomath*}...\end{linenomath*}

%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

>>>>

@@

@@

%% Keypoints, final entry on title page.
%% \begin{abstract} begins second page

@@

\begin{abstract}
Planetary science encompasses a broad number of research fields and
brings together several research communities (geologists, astronomers,
physicists, geochemists, etc.). Planetary missions produce an impressively growing 
amount of diverse data requiring an evolution from a mostly manual-based visual 
analysis to a more automated and quantitative analysis.
Interoperability, openness of data formats, and shared processing techniques are 
a necessity to efficiently extract scientific information from the data and 
to guarantee the reproducibility of the scientific results.
Unfortunately, the technologies and data formats used by researchers for
planetary surface studies and the field of astronomy 
diverged as these related, but almost completely isolated domains evolved.

In this paper we will describe how a small addition to the Flexible Image Transport
System (FITS) standard, widely used in the astronomy investigations, will allow FITS
to be more easily used in planetary surface investigations.
We will also show how FITS metadata can easily be transformed
in the PDS4 metadata archival model and lastly provide example implementations.
More than imposing a formal data model, a FITS description for planetary data aims
to simplify sharing data across the planetary and astronomy domains and promoting
interoperability from raw data formatting to final visualization.
\end{abstract}

@@

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition,
transfer and archiving of telescope data by astronomical observatories.
It had been used for data exchange and archiving from several orbital telescopes.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory (VO).
FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications so that FITS files can be embedded in PDS datasets.
Finally, it is supported by a large number of open libraries
and software tools.

@@

\begin{table}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: the corresponding PDS4 terminology is
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator: those three projections are not used for archival or analysis.}
\end{tabular}
\label{table:proj}
\end{table}

Once reprojected in one of the standard projections listed in Table \ref{table:proj}
Planetary images or maps can be described using the WCS scheme
\citep{greisencalaII,calagreisenI}.
In WCS context pixels are first translated in an intermediate pixel coordinate system
via linear transformations, then mapped on the grid specified by the \texttt{CTYPE}
keyword, containing the projection code and the coordinate system.
The coordinate system must be defined with respect to a specific inertial reference frame.
Multiple coordinate system descriptions are allowed in the same header: this is a
valuable property for planetary surface imagery, as angular (`deg') and linear (`m')
description of the surface are often needed at the same time.
Further below 3D keywords will be proposed for FITS metadata, in order to standardize
the conversion between angular and linear coordinates: alternate description will still
be important in our framework as standard libraries (e.g.
wcslib\footnote{http://www.atnf.csiro.au/people/mcalabre/WCS/wcslib/} or
astropy\footnote{http://www.astropy.org/}, \citet{astropy}) and some FITS visualization
software (e.g. ds9\footnote{http://ds9.si.edu/}) already implement them, making linear
coordinate tracking more efficient.

@@

\item{\texttt{DATE-OBS}.
It contains the date of the observation, provided as ISO 8601
string. The value shall be assumed to refer to the start of the observation,
unless another interpretation is clearly explained in the comment field.
It is assumed to be expressed in UTC but other systems are accepted as long as they
are specified in the ISO string\footnote{See table 30 in the Standard FITS document https://fits.gsfc.nasa.gov/standard40/fits\_standard40aa.pdf}. 
PDS4 provides a \texttt{Time\_Coordinates} class with attributes
\texttt{start\_date\_time} and \texttt{stop\_date\_time} providing start and
stop date and time appropriate to the product being labeled, always in UTC.
The \texttt{time\_min} and \texttt{time\_max} fields are mandatory in the EPN-TAP
scheme, they must always be formatted in UTC as a ISO string (planned in EPN-TAP v2.1).}

@@

@@

In addition to that, a QGIS plugin has been developed\footnote{https://github.com/epn-vespa/VO\_QGIS\_plugin}
\citep[][in preparation]{Minin} in order to download and display files exposed via EPN-TAP servers.

@@

\subsection{Aladin}
Aladin is both a sky atlas and a portal to access data -- mainly images and
catalogues -- available through the VO.
It is able to load FITS files and to display images taking into account various
WCS projections.
Like DS9, Aladin is popular and broadly used in the part of the astronomical
community manipulating images.
So far it is also the reference generator and visualizer of Hierarchical
Progressive Survey (HiPS).
HiPS\footnote{http://www.ivoa.net/documents/HiPS/} is a recommendation endorsed
by the International Virtual Observatory Alliance (IVOA).
It describes how to access to sky survey data stored at various spatial
resolutions, offering a progressive view of possibly very large datasets.
HiPS are usually generated from a collection of WCS tagged FITS files.
Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
been updated to support planetary conventions (see Figure \ref{fig:aladin}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendix goes here
%
% \appendix resets counters and redefines section heads
% but doesn't print anything.
% After typing \appendix
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Optional Glossary, Notation or Acronym section goes here:
%
%%%%%%%%%%%%%%
% Glossary is only allowed in Reviews of Geophysics
%\begin{glossary}
%\term{Term}
% Term Definition here
%\term{Term}
% Term Definition here
%\term{Term}
% Term Definition here
%\end{glossary}

@@

%\bibliography{ess_marmo}

@@

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%% ------------------------------------------------------------------------ %%
%
%  IN-TEXT LISTS
%
%% ------------------------------------------------------------------------ %%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
